---
title: 'Final Project - Step 2 (15 Points)'
subtitle: "PSTAT100: Data Science Concepts and Analysis" 
author:
    affiliations:
      - name: "Spring 2025"
affiliation-title: "Quarter"
format: 
 pdf: 
    latex_engine: xelatex
    geometry: "letterpaper, margin=0.3in"
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```

:::: callout
::: {style="text-align: center"}
[ **STUDENT NAMES** ]{style="color: blue;"}
:::

-   Luke Drushell (drushell)
-   David Pacheco (dpacheco)
-   Rafael Romero (romeroflores)
-   Kai Barker (kaibarker)
-   Diana Lee (dianadlee)
-   Davian Valencia (davianvalencia)
::::

::: callout-caution
## Due Date

The deadline for this step is **Friday, May 9, 2025**.
:::

::: callout-tip
# Instructions

In this step, you will develop clear research questions and hypotheses
based on your selected dataset, and conduct a thorough Exploratory Data
Analysis (EDA). This foundational work is crucial for guiding your
analysis in the following steps.
:::

# Step 2: Research Questions, Hypotheses, and Exploratory Data Analysis (EDA)

## Research Questions

**Question 1** How are genres linked to the vote_average of movies? And
does release year play a measurable role?

**Question 2** Are movies getting longer as time goes on? Does this
affect any aspect of the movie? (ex: ratings and revenue)

**Question 3** Does the budget of the movie affect how good it is. In
other words, is the average rating affected by the movies budget.

## Hypotheses

**Hypothesis 1** Movies are becoming more popular but less liked: As
year goes up popularity goes up proportionally, but vote average drops.
Maybe this is only in action/adventure genres, whereas
comedies/romances/dramas the opposite might be seen.

**Hypothesis 2** Despite the shorter attention span of our generation,
movies have been slowly increasing runtime, which has been seen to
increase the ratings and revenue of a movie.

**Hypothesis 3** While users rating or a movies quality is not always a
reflection of how much money was spent, a higher budget should afford
better actors, better production, etc and should lead to an overall
better movie in some aspects.

## Exploratory Data Analysis (EDA)

## Data Cleaning

Data came relatively clean, some columns such as indices, or long
descriptions are often missing so those columns will be directly
removed. Finally some rows have a few zero'd values, which will be delt
with on the case by case basis described in the comments.

```{r}
library(dplyr)
dataset <- read.csv("top_1000_popular_movies_tmdb.csv")
dataset <- dataset %>% select(-tagline)
dataset <- dataset %>% select(-X)
dataset <- dataset %>% select(-id)
dataset <- dataset %>% select(-overview)

#since half the data is missing this info we'll include it
revenue_zeros <- sum(dataset$revenue == 0, na.rm = TRUE)
budget_zeros  <- sum(dataset$budget == 0, na.rm = TRUE)
cat("Rows with revenue == 0:", revenue_zeros, "\n")
cat("Rows with budget  == 0:", budget_zeros,  "\n")
#------

#only 175 are missing runtimes though, so they can be purged
runtime_zeros <- sum(dataset$runtime == 0, na.rm = TRUE)
cat("Rows with runtime == 0:", runtime_zeros, "\n")
#------

#the dataset to be manipulated from here on out!
dataset <- dataset[!(dataset$runtime == 0), ]
```

## Descriptive Statistics

Some things that are interesting here: The mean average vote is 6.395,
with the 1st quartile being 6. It would be possible to filter true
vote_averages of 0 from movies with no votes by simply taking taking the
complement of when vote_count is 0, which may be something someone
mapping votes should consider. Another interesting thing here is how low
the popularity mean is -- many of the top 10,000 movies are not actually
that popular compared to true blockbusters as shown with the max (Fast
X). Finally, it is interesting to see how runtime cany vary: why is
there a movie with a runtime of 2 minutes for instance? It appears there
are a variety of short films in the top 10,000.

```{r}
dataset %>%
  summary()
```

## Data Visualization

Since the vote_average variable provides some of the most meaningful data, providing valuable insight into audiences feelings on the movie, we will make several visual plots so we can explore its distribution and relationship with other variables to better understand viewer preferences and trends.

First, we can explore the distribution of the vote average over all the movies in the dataset. As a reminder, the variable is the average rating of each movie, not the average number of votes. 

```{r}
#| echo: false
library(ggplot2)

ggplot(dataset, aes(x=vote_average)) +
  geom_histogram(bins=50, fill="maroon", color="black")+
  labs(title="Histogram of Vote Averages",
       x="Average Rating",
       y="Frequency") 
```
We can observe that the distribution of the vote averages looks approximately normal, which is further evidenced by the large sample size (~10,000) and the central limit theorem. Interestingly, there is a large cluster of movies with an average rating of 0.0. This could be due to relevancy of the movie, where few people have cared to rate it, lack of data due to the release date of the movie, genuine dislike of the movie, or some other error. Most likely it is due to an error or lack of data for the movies rating, since any actual value of 0.0 would be nearly impossible unless it lacked a single rating. As discussed previously, these could be omitted when exploring the relationship between vote average and other variables.

Now that we have an idea of how the vote average variable looks and how it is distributed, we can explore some of its relationships with other variables in the dataset. One relationship that could be interesting to explore is the average ratings by genre. We can model this relationship with a set of boxplots, each representing a genre, ordered by median average rating (as boxplots are centered about the median), with the boxes themselves modeling the middle 50\% of the data, and the whiskers indicating the spread of the non-boxed values.
```{r}
#| echo: false
library(tidyverse)
#first we need to get all the individual genres
#genres are like tags so some movies have multiple 
#genres (also removed vote avg of 0 as discussed before)
genre_sep <- dataset %>%
  mutate(genres = str_remove_all(genres, "\\[|\\]|'")) %>%
  separate_rows(genres, sep=",\\s*") %>%
  filter(genres != "", vote_average != 0) %>%
  group_by(median(vote_average))

#for some reason, only 2 rows in the seperated dataset were NA
genre_sep <- na.omit(genre_sep)

#want to reorder so that we can can easier tell the smallest to largest median values
genre_sep <- genre_sep %>%
  mutate(genres = fct_reorder(genres, vote_average, .fun = median, .desc = TRUE))

ggplot(genre_sep, aes(x=vote_average, y=genres, fill = genres)) +
  geom_boxplot() +
  labs(title="Boxplots for Average Rating by Genre", 
       x="Average Rating",
       y="Genre") +
  theme(legend.position = "none")
```
From our boxplots, we can observe that the smallest median vote average comes from Horror movies, and the largest from Documentaries. While this may make more or less sense depending on personal opinions, Horror movies and Documentaries might tend to have more polarizing opinions, where documentaries might appeal to larger audiences, or audiences that were already interested in the movie, depending on the subject covered, and Horror movies might receive less praise if people are turned off by their gory nature. 

Another relationship that could be important to explore is that between the vote average and budget of the movies. While budget should not inherently increase the quality of the movie, it should eliminate some aspects that could lower ratings, such as production, camera quality, visual effects, and the level of the cast afforded. We can model this relationship with a scatter plot
```{r}
#| echo: false
##filter vote avg of 0 out
data_wo_zeros <- dataset %>%
  filter(vote_average != 0, budget != 0)

ggplot(data_wo_zeros, aes(x=vote_average, y=budget/(1000000))) +
  geom_point(alpha=0.075) +
  labs(title="Average Rating vs. Movie Budget",
       x="Average Rating",
       y="Budget (Millions)")
```
Values in the scatterplot have had their opacity turned down to better highlight large clusters of data. The scatterplot might not be the best method of analyzing this relationship as the vast number of datapoints leads to more clutter. However, we can observe that a majority of movies, regardless of budget seem to fall between an average rating of 5-7.5. Further, most of the movies appear to have a budget of about 50 million or less. We can also observe that the movie with the largest budget, around 450 million, does not have the highest average rating, and there is a large amount of movies with a low budget that are mixed in with the cluster of the 5-7.5 vote average. It would appear that the movies with the highest average rating have a budget of around 100 million dollars. To make a better conclusion about the relationship between budget and ratings, further statistical analysis would be warranted.

We can also investigate Budget vs. Revenue with a scatterplot. Due to the wide range of monetary values, we can apply log scales to both axes to help us visualize the pattern more clearly.

```{r}
ggplot(dataset, aes(x = budget, y = revenue, color = vote_average)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c() +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Budget vs Revenue (Colored by Vote Average)")
```

From the plot above, we can observe a clear positive correlation between budget and revenue, suggesting that movies with a higher budget tend to bring in higher revenues. However, the spread also reveals significant variability as some lower-budget films achieve substantial revenues, and some high-budget films underperform. The color gradient adds another layer of insight. While many high-revenue films also have moderate-to-high vote averages, there are notable exceptions: some well-rated movies (darker color) exist across all budget levels, indicating that critical reception (vote average) is not solely dependent on budget size.

Furthermore, this visual complements our exploration of revenue and ratings trends and emphasizes the complexity in predicting movie success based on production costs alone.



We can also investigate Popularity vs. Average Rating with a scatter plot. This allows us to explore whether popular movies tend to receive higher or lower ratings.

```{r}
ggplot(dataset, aes(x = popularity, y = vote_average)) +
  geom_point(alpha = 0.6, color = "blue") +
  labs(
    title = "Relationship between Movie Popularity and Average Rating",
    x = "Popularity Score",
    y = "Average Rating (Vote Average)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12)
  )
```

From the plot above, we can see that there is a wide distribution of average ratings across all popularity levels. This suggests that popularity is not directly tied to higher ratings. Highly popular movies can have a wide range of ratings, from very low to very high. Some low-popularity movies have high ratings, indicating that they may be critically acclaimed but less known to the general audience. The scatter plot does not show a clear linear relationship between popularity and rating, indicating that factors other than popularity may significantly influence ratings. This visual complements our analysis of other factors like budget and run time and further highlights the complexity of predicting a movie’s quality based solely on its popularity.


Another relationship we can investigate is weather the number of votes(vote count) and average rating(vote_average) have a correlation.This can help us understand if heavily voted movies tend to be perceived better or worse by the general public.
```{r}
# Filter out movies with 0 votes or 0 ratings
vote_data <- dataset %>%
  filter(vote_count > 0, vote_average > 0)

ggplot(vote_data, aes(x = vote_count, y = vote_average)) +
  geom_point(alpha = 0.3, color = "purple") +
    scale_x_log10() +
  labs(
    title = "Relationship Between Vote Count and Average Rating",
    x = "Number of Votes (log scale)",
    y = "Average Rating"
  ) +
  theme_minimal()
```

This scatter plot analyzes the correlation between the number of votes and the average rating given by votes. There are very few highly voted movies with ratings below 5 or above 9, which would be expected because of the central limit effect. There is not a strong linear relationship between the variables but it is important to see that the movies with a higher number of votes tend to land more in the moderate area when it comes to their average ratings and this is a result of the rating being drawn from more opinions.

```{r}
#| echo: false
library(ggwordcloud)

#split the genres inside the vectors in each row
genresplit <- dataset %>% 
  separate_rows(genres, sep = "[\\[\\]',\\s]+") %>% filter(genres != "") %>% count(genres, sort = TRUE)  

#make the word cloud
ggplot(genresplit, aes(label = genres, size = n)) +
  geom_text_wordcloud(area_corr = TRUE) +
  scale_size_area(max_size = 20) +
  theme_minimal()
```

Using the word cloud above, we can visually see which genre is most popular within this data set. We can see that "Drama" is the most common movie genre, while genres, such as "Documentary" and "Western", seem as though they are least common. This tells us that within the top 1000 movies, "Drama" is the most popular. This knowledge can be then used to further analyze other variables in relation to their movie genre.
